Implications of prey skill and antipredator behaviour in the acquisition of predator expertise: a virtual ecological study

### 2025-08-19 manuscript/manuscript.Rmd
updated intro and results adding space covered

dat <- fread(
  file.path(
    path, "FraserFrancoetal2023-data.csv"
  ),
  select = c(
    "timestamp", "match_id", "realm_id",
    "environment_id", "avatar_id", "predator_id",
    "game_duration", "total_xp_pred", "cumul_xp_pred",
    "hunting_success", "pred_speed", "prey_avg_speed", 
    "prey_avg_rank"
  )
)

@@

@@

Predation is a fundamental biological process acting as an agent of evolutionary change that regulates prey populations, limits the spread of diseases, and mediates energy flow across trophic levels [@hairston1993;@abrams2000;@ripple.etal2014;@wirsing.etal2021]. These processes are driven by changes in predation risk and prey selection across landscapes as predators select suitable habitats to hunt and adjust their strategies accordingly  [@quevedo.etal2009;@pettorelli.etal2015;@labarge.etal2024;@schmitz2017]. Central theory suggests that predators become more efficient hunters by acquiring expertise through practicing and learning the proper skills and tactics to locate, select, and capture their prey [@woo.etal2008;@wooster.etal2023;@labarge.etal2024]. This raises the hypothesis that differences in hunting success between predators may be attributed to differences in the capacity to acquire hunting expertise. Hence, unraveling how this expertise emerges remains a missing, but essential, link to assess how foraging behaviour shapes prey consumption during predator-prey interactions.

@@

Prey respond to predators by using antipredator tactics such as camouflage to avoid detection and rapid escapes to evade capture [@walker.etal2005;@kelley.magurran2011;@herbert-read.etal2017]. These strategies are hypothesized to drive differences in prey consumption among predators by disrupting their capacity to acquire hunting expertise [@wooster.etal2023]. This has been supported by experimental studies showing that certain camouflage tactics of prey can impair expertise acquisition in humans and birds [@martin.etal2012;@troscianko.etal2013]. For example, using a virtual experiment, @troscianko.etal2018 found in human subjects that disruptive colouration interfered with their ability to form search images, hindering improvements in detection times over repeated attempts. Given these observations, antipredator behaviour may also hold the potential to hinder the acquisition of predator expertise. For instance, some predators may be limited in their capacity to develop the necessary attributes (e.g. physical, physiological, neurological) for fast-paced hunting, which could impair their acquisition of hunting expertise if they hunt faster prey. Prey antipredator behaviour can also improve with experience through continued exposure to predators [@turner2006;@kelley.magurran2011;@lonnstedt.etal2012]. Indeed, animal reintroduction programs show that individuals trained to recognize predators display more antipredator behaviours and are 1.5 times more likely to survive than individuals that were never exposed to predators [@tetzalf.etal2019]. To our knowledge, there is little empirical evidence showing links between prey skill, antipredator behaviour, and the acquisition of predator expertise in human and nonhuman predators, representing a significant gap in our understanding of predator-prey interactions.

A recurring challenge impeding research on predator-prey behavioural interactions is the need to collect data simultaneously on both the predator and the prey. Here we mitigate these challenges by using *Dead by Daylight* as our study system, a videogame where four prey players must forage for resources while avoiding predation by a fifth player. Similar to agent-based simulations, *Dead by Daylight* provides controlled virtual environments to test ecological hypotheses [see @montiglio.etal2025 for a review], with the advantage of having real players that interact in the virtual space. In this game, the predator population comprises individuals that either ambush or hunt at high speeds (i.e., mean movement speed along an ambush-cursorial continuum of tactics), and their success is driven by the movement of the prey [@fraserfranco.etal2022]. The prey can increase their chances of survival by cooperating and moving fast to escape the predator [@cere.etal2021;@fraserfranco.etal2022;@santostefano.etal2024]. The game also elicits natural reactions in players such as freezing when predation is imminent (M.F.F., personal observations), corroborating virtual ecological studies showing that predation drives individual variation in risk perception [@beauchamp2020]. These observations outline how ecological phenomena can emerge from human interactions in virtual systems with fixed rules [@brosnan.etal2017;@kasumovic.etal2017]. Videogames also generate large volumes of data on thousands of interacting players throughout their lifetime in the game under realistic, controlled, and repeatable ecological scenarios. Hence, *Dead by Daylight* allows us to tackle fundamental questions about the role of antipredator behaviour and experience in predator-prey interactions.

In this study, we assess how predator expertise emerges over repeated encounters with prey using data from players in *Dead by Daylight*. We quantify expertise acquisition as the relationship between hunting success (i.e., prey capture) and repeated experience (i.e., cumulated matches). First, we test the hypothesis that predator success will increase with experience up to some point at which it will stabilize [@dukas2019]. However, we hypothesize that prey speed will influence predator expertise acquisition by modulating the rate of gain in expertise at the population level. Second, we test the hypothesis that prey with more skill should have improved antipredator behaviour, and predict that prey with greater skills will move at faster speeds. Third, if prey speed influences hunting success at the population level, then we expect that the acquisition of predator expertise will vary among individuals.

@@

*Dead by Daylight* is a survival asymmetric (i.e.,, gameplay mechanics differ between two groups) multiplayer online game  developed by Behaviour Interactive Inc., in which players can play either as predators or prey. Every match includes only one predator and four prey. The objective of the predator is to hunt and capture the prey, and the objective of the prey is to search for resources while avoiding the predator. The resources are in the form of power generators that, once all activated, will enable the prey to escape through one of two exit doors. A skill-based matchmaking algorithm determines the composition of the predator and prey group in a match. A match ends when the predator kills all the prey available (i.e., that have not escaped), or when the last remaining prey escapes the virtual environment.

@@

Our study population comprised `r toString(format(length(unique(dat$predator_id))))` players that played at least 300 matches. We monitored all their matches from the first to a maximum of 500 matches, with a total record of `r toString(formatC(length(unique(dat$match_id)), big.mark = " "))` matches overall. The predator-players' experience in our population sample varied between `r toString(format(min(dat$total_xp_pred)))` and `r toString(format(max(dat$total_xp_pred)))` matches played. These matches lasted between `r toString(format(round(min(dat$game_duration / 60))), digits = 2)` and `r toString(format(round(max(dat$game_duration / 60))), digits = 2)` min (mean = `r toString(format(round(mean(dat$game_duration / 60))), digits = 2)` min). We recognize that we could have introduced a bias by retaining only those individuals, as they might already be seasoned video game enthusiasts and exhibit expert-level performance in their early matches in *Dead by Daylight*. Thus, we tested sample bias by comparing a random sample of players that played either 20 to 50 matches, 51 to 100 matches, or 101 to 300 matches during the same timeframe as our population sample. We then took the first 20 matches played by these players, including those from our population sample, and compared their mean hunting success using a Bayesian hierarchical linear model. We found that all four groups had similar success as predators in their first 20 matches (Appendix 1: Table S1 and Figure S1), giving us confidence that we did not introduce a sampling bias.

We collected the following information for every match in our population sample: the player's anonymous ID, the predator player's hunting success, the predator player's cumulated experience, and the mean rank and speed of the group of prey that the predator player encountered. We defined hunting success as the number of prey consumed during the match (min = `r toString(format(min(dat$hunting_success)))`, max = `r toString(format(max(dat$hunting_success)))`). We defined the predator's cumulative experience as the number of matches played as the predator prior to the match being monitored. For example, the first match of a player would have a cumulative experience value of 0, while the tenth match would have a value of 9. Laslty, we measured the prey's rank as the mean rank of the four individual prey in a match (mean = `r toString(format(round(mean(dat$prey_avg_rank, na.rm = T), digits = 2), nsmall = 2))` ± `r toString(format(round(sd(dat$prey_avg_rank, na.rm = T), digits = 2), nsmall = 2))`), and the prey's speed as the mean travel speed of the four individual prey in a match (mean = `r toString(format(round(mean(dat$prey_avg_speed, na.rm = T), digits = 2), nsmall = 2))` ± `r toString(format(round(sd(dat$prey_avg_speed, na.rm = T), digits = 2), nsmall = 2))` m/s).

We used the mean rank of the prey as a proxy of their skill. Player ranks vary between `r toString(format(min(dat$prey_avg_rank)))` and `r toString(format(max(dat$prey_avg_rank)))`, with a rank of 1 indicating the highest skill. The ranking system in *Dead by Daylight* was implemented by the company to pair players in a match based on their skill (https://deadbydaylight.fandom.com/wiki/Rank). The skill of a player increases based on their performance from match to match. While the ranking system represents an approximation of prey skill, it was the most readily available metric we could access to determine the skill of the prey group, as we did not have access to the prior matches of the prey. In addition, the pairing system is subject to variation depending on factors such as player availability, which can result in predator-prey groups with unbalanced skill. This allowed us to evaluate the effect of the prey's skill on predator success.

@@

We tested how predators developed their expertise by computing five Bayesian generalized additive mixed models (GAMM) with thin-plate regression splines, all of which estimated the relationship between hunting success and the predators' cumulative experience. We parametrised the models following the method of @pedersen.etal2019. For the first and second models (I and II), we assume that individual predators share a similar relationship between success and experience, but that this relationship can vary among them (e.g. predator 1 has a steeper curve than predator 2). This enabled us to test whether predators differed in the development of their expertise. Both models included a common global smoothing function (i.e., population effects) and random smooths for the predator ID (i.e., individual effects). Model I included the standardized match duration as a covariate, while model II included the standardized match duration and the average rank of the prey group, allowing us to test if the variation among predators in expertise acquisition was driven by prey skill. The third model had the same structure as model II but without a global smoother. This model assumes that predators do not share a common relationship between success and experience. In this model, we control for the standardized match duration and the average rank of the prey group. The fourth (IV) and fifth (V) models were expansions of model II, but included the standardized prey speed to assess its effect on the relationship between success and experience. Model IV does not have a standard global smoother, while mother V includes one.

We configured the five models to use a modified beta-binomial distribution. Hunting success was modeled as the number of prey consumed out of four, with the probability of success estimated from a Beta distribution ($Beta(\mu_i, \phi)$) characterized by a mean ($\mu \in [0,1]$) and a precision ($\phi > 0$) parameter. We used a logit link function to estimate $\mu_i$ where $\mu_i = \frac{exp(\eta_i)}{exp(\eta_i) + 1}$ and $\eta_i$ is the linear predictor. The precision parameter ($\phi$) was estimated using a log link function. We used ten basis functions (K = 10) for the models to estimate the relationship between hunting success and experience. We assumed that the random intercepts for the predator ID ($id$) followed a Gaussian distribution with estimated standard deviation ($id \sim N(0,\ I\ \sigma_{id}^{2})$). We compared the predictive accuracy of all five models using approximate leave-one-out cross-validation with Pareto-smoothed importance sampling [@vehtari.etal2017;@piironen.vehtari2017;@vehtari.etal2022].

We defined weakly informative Gaussian priors for the intercept ($N(0,\ 0.5^{2})$) and the global trend of cumulative experience ($N(0,\ 2^{2}))$). Following @fraserfranco.etal2022, we defined a positive Gaussian prior ($N(1,\ 0.5^{2}))$) for the slope of the game duration because longer trials lead to greater success, a Gaussian prior ($N(0,\ 1^{2}))$) for the slope of the prey rank, and a negative Gaussian prior on prey speed ($N(-1,\ 0.5^{2})$) because encountering faster prey is associated with lower success in this system [@fraserfranco.etal2022]. We defined a positive Gaussian prior on the precision parameter ($N(2,\ 0.5^{2})$). We employed weakly informative half-Gaussian priors on all the standard deviation parameters ($N(0,\ 0.5^{2})$).

To test the hypothesis that prey speed increases with prey skill, we fitted a Bayesian linear regression estimating the relationship between prey speed and the standardized prey rank. We controlled for the game duration, using it as a linear covariate in the model, and included the predator identity as a random effect. We used a weakly informative Gaussian prior on the slope of the game duration ($N(0,\ 0.5^{2})$) and a weakly informative negative Gaussian prior on the slope of the prey rank ($N(-0.5,\ 1^{2})$) because prey of greater skill (i.e., rank 1 is the highest skill) should move faster. Lastly, we used a weakly informative Gaussian prior on the intercept ($N(2,\ 0.5^{2})$) because prey should move at around two meters per second, and half-Gaussian priors on the standard deviation parameters ($N(0,\ 0.5^{2})$).

@@

We parametrised the GAMMs to run four chains of 1500 iterations each, discarding the first 500 iterations of each chain as warm-up, and sampling one parameter value per iteration thereafter. Similarly, the linear model ran four chains of 1500 iterations each, sampling a parameter value at every iteration, and discarding the first 500 iterations. We thus obtained 4000 posterior samples per parameter in each model. We assessed the convergence of the chains using trace plots, R-hat diagnostics with a threshold set between 1.00 and 1.02, and effective sample sizes (ESS) with a threshold of >100 [@vehtari.etal2021]. We also performed posterior predictive checks which showed an adequate fit of the models.

@@

We tested the hypothesis that antipredator behaviour impairs the acquisition of predator expertise at the population level by comparing the global trends of model I (only controlling for game duration), model II (controlling for game duration and prey rank) and model V (controlling game duration, prey rank, and prey speed) using the finite differences method (Figure 1). This allowed us to locate the point at which hunting success was optimized. At the individual level, we tested the hypothesis that antipredator behaviour generate differences among predators in expertise acquisition by comparing the individual-level variance parameters. Specifically, we compared the standard deviations of 1) the random intercepts (i.e., mean differences in hunting success), 2) the random slopes (i.e., linear component relating hunting success with experience), and 3) the curve wiggliness (i.e., nonlinear component relating hunting success with experience).

@@

# Add model column
loo_tab[
  , model1 := c(
    "(V) predator xp + ID smoothers + prey rank + prey speed",
    "(IV) ID smoothers + prey rank + prey speed",
    "(II) predator xp + ID smoothers + prey rank",
    "(III) ID smoothers + prey rank",
    "(I) predator xp + ID smoothers"
  )
]

@@

  # Table caption
  set_caption(
    caption = "Table 1. Leave-one-out cross-validation results comparing the predictive performance of the five GAMMs estimating the relationship between hunting success and predator experience. All models reported include the game duration as a control. Only differences in smoothing functions and prey covariates are reported.",
    style = "Table Caption"
  ) %>%

@@

Our analyses also revealed important differences among individual predators in the acquisition of expertise, indicating that predators varied in their capacity to improve over time. However, prey speed did not influence among-individual differences in expertise. The large indivual variation in expertise acquisition emerged when we controlled for the prey skill and persisted when we controlled for the prey's speed, suggesting that it would be observed even if predators hunted prey of average speed. This suggests that other proximal or ecological factors may be at play. For example, predators have limited attention spans that restrict diet choice and the formation of search images when they need to focus on hunting multiple prey simultaneously [@dukas.etal2001;@troscianko.etal2018]. Chasing prey demands specialized cognitive abilities and coordination that are energetically costly [@kelley.magurran2011]. Hence, individuals with neurological or physical attributes allowing them to better concentrate on the hunting task may have shown greater improvement in success over time [@dukas2019]. Alternatively, predators may also shape their own expertise via their hunting tactic. For example, faster hunters may push their prey to move faster and display altruistic behaviours, which, in turn, may hinder their own expertise acquisition due to the increased difficulty of hunting such prey [@fraserfranco.etal2022;@cere.etal2024]. Lastly, longer time intervals between hunting events are hypothesized to hinder or delay the acquisition of expertise due to information loss [@endler1991;@wright.etal2022;@wooster.etal2023]. For example, a predator that played 300 matches within six months might forget critical information related to prey detection or escape patterns compared to one that played 300 matches within six weeks. Therefore, investigating the impact of the aforementioned elements may reveal important insights into the outcome of predator-prey interactions, but integrating them into a unified framework remains challenging due to computational constraints and limitations in current modeling approaches.

>>>>

dat <- fread(
  file.path(
    path, "FraserFrancoetal2025-data.csv"
  ),
  select = c(
    "timestamp", "match_id", "realm_id",
    "environment_id", "avatar_id", "predator_id",
    "game_duration", "total_xp_pred", "cumul_xp_pred",
    "hunting_success", "pred_speed", "prey_avg_speed", 
    "prey_avg_rank", "prey_avg_amount_tiles_visited"
  )
)

@@

**à compléter**

@@

Predation is an agent of evolutionary change that regulates prey populations, limits the spread of diseases, and mediates energy flow across trophic levels [@hairston1993;@abrams2000;@ripple.etal2014;@wirsing.etal2021]. These processes are driven by changes in predation risk and prey capture across landscapes as predators select suitable habitats to hunt and adjust their strategies accordingly  [@quevedo.etal2009;@pettorelli.etal2015;@labarge.etal2024;@schmitz2017]. Theory suggests that predators become more efficient hunters by acquiring expertise through practicing and learning the proper skills and tactics to locate, select, and capture their prey [@woo.etal2008;@wooster.etal2023;@labarge.etal2024]. This raises the hypothesis that differences in hunting success between predators may be attributed to differences in the capacity to acquire hunting expertise. Hence, unraveling how expertise emerges is essential to better assess how foraging behaviour shapes prey consumption during predator-prey interactions.

@@

Prey respond to predators by using antipredator tactics such as camouflage to avoid detection and rapid escapes to evade capture [@walker.etal2005;@kelley.magurran2011;@herbert-read.etal2017]. These strategies are hypothesized to drive differences in prey consumption among predators by disrupting their capacity to acquire hunting expertise [@wooster.etal2023]. This has been supported by experimental studies showing that prey camouflage can impair expertise acquisition in humans and birds [@martin.etal2012;@troscianko.etal2013]. For example, using a virtual experiment, @troscianko.etal2018 found in human subjects that disruptive colouration interfered with their ability to form search images, hindering improvements in detection times over repeated attempts. Beyond prey camouflage, there is limited evidence to suggest that antipredator behaviour hinders the acquisition of predator expertise. Yet, some predators may be limited in their capacity to develop the necessary attributes (e.g. physical, physiological, neurological) for fast-paced hunting, which could impair their acquisition of hunting expertise when facing faster prey. Prey antipredator behaviour can also improve through continued exposure to predators [@turner2006;@kelley.magurran2011;@lonnstedt.etal2012]. For instance, prey alter their activity patterns under increasing predation risk [@wirsing2021]. Animal reintroduction programs show that individuals trained to recognize predators display more antipredator behaviours and are 1.5 times more likely to survive than individuals that were never exposed to predators [@tetzalf.etal2019]. Hence, encountering prey with greater skill or experience may also impair expertise acquisition in predators. To our knowledge, the relationships among prey skill, antipredator behaviour, and the acquisition of expertise by human and nonhuman hunters remain unclear, representing a significant gap in our understanding of predator-prey interactions.

A recurring challenge impeding research on predator-prey behavioural interactions is the need to collect data simultaneously on both the predator and the prey over successive interactions. Here we mitigate these challenges by using *Dead by Daylight* as our study system, a videogame where four prey players must forage for resources while avoiding predation by a fifth player. Similar to agent-based simulations, *Dead by Daylight* provides controlled virtual environments to test ecological hypotheses [see @montiglio.etal2025 for a review], with the advantage of having real players that interact in the virtual space. In this game, the predator population comprises individuals that either ambush or hunt at high speeds (i.e., mean movement speed along an ambush-cursorial continuum of tactics), and their success is driven by the movement of the prey [@fraserfranco.etal2022]. The prey can increase their chances of survival by cooperating and moving fast to escape the predator [@cere.etal2021;@fraserfranco.etal2022;@santostefano.etal2024]. The game also elicits natural reactions in players such as freezing when predation is imminent (M.F.F., personal observations), corroborating virtual ecological studies showing that predation drives individual variation in risk perception [@beauchamp2020]. These observations outline how ecological phenomena can emerge from human interactions in virtual systems with fixed rules [@brosnan.etal2017;@kasumovic.etal2017]. Videogames also generate large volumes of data on thousands of interacting players throughout their lifetime in the game under realistic, controlled, and repeatable ecological scenarios. Hence, *Dead by Daylight* allows us to tackle fundamental questions about the role of antipredator behaviour and experience in predator-prey interactions.

In this study, we assess how predator expertise emerges over repeated encounters with prey using data from players in *Dead by Daylight*. We quantify expertise acquisition as the relationship between hunting success (i.e., prey capture) and repeated experience (i.e., cumulated matches). First, we test the hypothesis that predator success will increase with experience up to some point at which it will stabilize [@dukas2019]. However, we hypothesize that prey speed and space coverage will influence predator expertise acquisition by modulating the rate of gain in expertise at the population level. Second, we test the hypothesis that prey with more skill should have improved antipredator behaviour, and we predict that prey with greater predator-avoidance skills will move at faster speeds and cover more space in the virtual environment. Third, if prey antipredator behavior influences the hunting success of predators at the population level, then we expect that the acquisition of predator expertise will vary among individuals.

@@

*Dead by Daylight* is a survival asymmetric (i.e., gameplay mechanics differ between two groups) multiplayer online game developed by Behaviour Interactive Inc. (Montreal, Canada), in which players can play either as a predator or prey. Every match includes only one predator and four prey. The objective of the predator is to hunt and capture the prey, and the objective of the prey is to search for resources while avoiding the predator. The resources are in the form of power generators that, once all activated, will enable the prey to escape through one of two exit doors. A skill-based matchmaking algorithm determines the composition of the predator and prey group in a match. A match ends when the predator kills all the prey available (i.e., that have not escaped), or when the last remaining prey escapes the virtual environment.

@@

Our study population comprised `r toString(format(length(unique(dat$predator_id))))` players who played at least 300 matches. We monitored all their matches from the first to a maximum of 500 matches, with a total record of `r toString(formatC(length(unique(dat$match_id)), big.mark = ","))` matches overall. The predator-players' experience in our population sample varied between `r toString(format(min(dat$total_xp_pred)))` and `r toString(format(max(dat$total_xp_pred)))` matches played. These matches lasted between `r toString(format(round(min(dat$game_duration / 60))), digits = 2)` and `r toString(format(round(max(dat$game_duration / 60))), digits = 2)` min (mean = `r toString(format(round(mean(dat$game_duration / 60))), digits = 2)` min). We recognize that we could have introduced a bias by retaining only those individuals, as they might already be seasoned video game enthusiasts and exhibit expert-level performance in their early matches in *Dead by Daylight*. Thus, we tested sample bias by comparing the results from our dataset to those from a random sample of players who played either 20 to 50 matches, 51 to 100 matches, or 101 to 300 matches during the same timeframe. We then took the first 20 matches played by these players, including those from our population sample, and compared their mean hunting success using a Bayesian hierarchical linear model. We found that all three groups of players had very similar success as the predator in their first 20 matches among them and compared to those used as our focal sample (Appendix 1: Table S1 and Figure S1), giving us confidence that we did not introduce a significant sampling bias.

We collected the following information for every match in our population sample: the player's anonymous ID, the predator player's hunting success, the predator player's cumulated experience, and the mean rank, speed, and space coverage of the group of prey that the predator player encountered. We defined hunting success as the number of prey consumed during the match (min = `r toString(format(min(dat$hunting_success)))`, max = `r toString(format(max(dat$hunting_success)))`). We defined the predator's cumulative experience as the number of matches played as the predator prior to the match being monitored. For example, the first match of a player would have a cumulative experience value of 0, while the tenth match would have a value of 9.

We used the mean rank of the prey as a proxy of their predator-avoidance skill. Player ranks vary between `r toString(format(min(dat$prey_avg_rank)))` and `r toString(format(max(dat$prey_avg_rank)))`, with a rank of 1 indicating the highest skill. The ranking system in *Dead by Daylight* was implemented by the company to pair players in a match based on their skill (https://deadbydaylight.fandom.com/wiki/Rank). The skill of a player increases based on their performance from match to match. While the ranking system represents an approximation of prey skill, it was the most readily available metric we could access to determine the skill of the prey group, as we did not have access to the prior matches of the prey. In addition, the pairing system is subject to variation depending on factors such as player availability, which can result in predator-prey groups with unbalanced skill, allowing us to evaluate the effect of the prey's skill on predator success. For a given match, we measured the prey's rank as the mean rank of the four individual prey (mean = `r toString(format(round(mean(dat$prey_avg_rank, na.rm = T), digits = 2), nsmall = 2))` ± `r toString(format(round(sd(dat$prey_avg_rank, na.rm = T), digits = 2), nsmall = 2))`), and the prey's speed as the mean travel speed of the four individual prey in a match (mean = `r toString(format(round(mean(dat$prey_avg_speed, na.rm = T), digits = 2), nsmall = 2))` ± `r toString(format(round(sd(dat$prey_avg_speed, na.rm = T), digits = 2), nsmall = 2))` m/s). Lastly, the prey space coverage represents the number of unique 16x16 tiles entered by the predator in a grid that covers the whole virtual environment (mean = `r toString(format(round(mean(dat$prey_avg_amount_tiles_visited, na.rm = T), digits = 2), nsmall = 2))` ± `r toString(format(round(sd(dat$prey_avg_amount_tiles_visited, na.rm = T), digits = 2), nsmall = 2))` tile visits).

@@

We tested how predators developed their expertise by computing six Bayesian generalized additive mixed models (GAMM) with thin-plate regression splines, all of which estimated the relationship between hunting success and the predators' cumulative experience. We parametrised the models following the method of @pedersen.etal2019. For the first and second models (I and II), we assume that individual predators share a similar relationship between success and experience, but that this relationship can vary among them (e.g. predator 1 has a steeper curve than predator 2). This enabled us to test whether predators differed in the development of their expertise. Both models included a common global smoothing function (i.e., population effects) and random smooths for the predator ID (i.e., individual effects). Model I included the standardized match duration as a covariate, while model II included the standardized match duration and the average rank of the prey group, allowing us to test if the variation among predators in expertise acquisition was driven by prey skill. The third model had the same structure as model II but without a global smoother. This model assumes that predators do not share a common relationship between success and experience. In this model, we control for the standardized match duration and the average rank of the prey group. The fourth (IV) and fifth (V) models were expansions of model II, but included the standardized prey speed to assess its effect on the relationship between success and experience. Model IV does not have a standard global smoother, while model V includes one. Lastly, model VI was an extension of model V and included all the aforementioned variables with the addition of the standardized space covered by the prey.

We configured the six models to use a modified beta-binomial distribution. Hunting success was modeled as the number of prey consumed out of four, with the probability of success estimated from a Beta distribution ($Beta(\mu_i, \phi)$) characterized by a mean ($\mu \in [0,1]$) and a precision ($\phi > 0$) parameter. We used a logit link function to estimate $\mu_i$ where $\mu_i = \frac{exp(\eta_i)}{exp(\eta_i) + 1}$ and $\eta_i$ is the linear predictor. The precision parameter ($\phi$) was estimated using a log link function. We used ten basis functions (K = 10) for the models to estimate the relationship between hunting success and experience. We assumed that the random intercepts for the predator ID ($id$) followed a Gaussian distribution with estimated standard deviation ($id \sim N(0,\ I\ \sigma_{id}^{2})$). We compared the predictive accuracy of all six models using approximate leave-one-out cross-validation with Pareto-smoothed importance sampling [@vehtari.etal2017;@piironen.vehtari2017;@vehtari.etal2022].

We defined weakly informative Gaussian priors for the intercept ($N(0,\ 0.5^{2})$) and the global trend of cumulative experience ($N(0,\ 2^{2}))$). Following @fraserfranco.etal2022, we defined a positive Gaussian prior ($N(1,\ 0.5^{2}))$) for the slope of the game duration, a Gaussian prior ($N(0,\ 1^{2}))$) for the slope of the prey rank, and negative Gaussian priors on prey speed and space coverage ($N(-1,\ 0.5^{2})$). We defined a positive Gaussian prior on the precision parameter ($N(2,\ 0.5^{2})$). We employed weakly informative half-Gaussian priors on all the standard deviation parameters ($N(0,\ 0.5^{2})$).

To test the hypothesis that prey speed and space coverage increases with prey skill, we fitted Bayesian linear regression models estimating the relationship between 1) prey speed and the standardized prey rank and 2) prey space coverage and the standardized prey rank. In both models, we controlled for the game duration, using it as a linear covariate, and included the predator identity as a random effect. We used weakly informative Gaussian priors on the slopes of the game duration ($N(0,\ 0.5^{2})$) and weakly informative negative Gaussian priors on the slopes of the prey rank ($N(-0.5,\ 1^{2})$) because prey of greater skill (i.e., rank 1 is the highest skill) should move faster and cover more space in the environment. Lastly, we used weakly informative Gaussian priors on the intercepts of prey speed ($N(2,\ 0.5^{2})$) and space coverage ($N(15,\ 0.5^{2})$) as we assumed that prey should move at around two meters per second and cover at least 15 tiles in the virtual environment. Lastly, we employed half-Gaussian priors on the standard deviation parameters ($N(0,\ 0.5^{2})$) in both models.

@@

We parametrised the GAMMs to run four chains of 1500 iterations each, discarding the first 500 iterations of each chain as warm-up, and sampling one parameter value per iteration thereafter. Similarly, the linear models ran four chains of 1500 iterations each, sampling a parameter value at every iteration, and discarding the first 500 iterations. We thus obtained 4000 posterior samples per parameter in each model. We assessed the convergence of the chains using trace plots, R-hat diagnostics with a threshold set between 1.00 and 1.02, and effective sample sizes (ESS) with a threshold of >100 [@vehtari.etal2021]. We also performed posterior predictive checks, which showed an adequate fit of the models.

@@

We tested the hypothesis that antipredator behaviour impairs the acquisition of predator expertise at the population level by comparing the global trends of model I (only controlling for game duration), model II (controlling for game duration and prey rank), model V (controlling for game duration, prey rank, and prey speed), and model VI (controlling for game duration, prey rank, prey speed, and prey space coverage) using the finite differences method (Figure 1). This allowed us to locate the point at which hunting success was optimized. At the individual level, we tested the hypothesis that antipredator behaviour generates differences among predators in expertise acquisition by comparing the individual-level variance parameters. Specifically, we compared the posterior distributions of standard deviations for 1) the random intercepts (i.e., mean differences in hunting success), 2) the random slopes (i.e., linear component relating hunting success with experience), and 3) the curve wiggliness (i.e., nonlinear component relating hunting success with experience).

@@

# Add model column
loo_tab[
  , model1 := c(
    "(VI) predator xp + ID smoothers + prey rank + prey speed",
    "(V) predator xp + ID smoothers + prey rank + prey speed",
    "(IV) ID smoothers + prey rank + prey speed",
    "(II) predator xp + ID smoothers + prey rank",
    "(III) ID smoothers + prey rank",
    "(I) predator xp + ID smoothers"
  )
]

@@

  # Table caption
  set_caption(
    caption = "Table 1. Leave-one-out (loo) cross-validation results comparing the predictive performance of the five GAMMs estimating the relationship between hunting success and predator experience. All models reported include the game duration as a control. Only differences in smoothing functions and prey covariates are reported.",
    style = "Table Caption"
  ) %>%

@@

Our analyses also revealed important differences among individual predators in the acquisition of expertise, indicating that predators varied in their capacity to improve over time. However, prey speed did not influence among-individual differences in expertise. The large indivual variation in expertise acquisition emerged when we controlled for the prey skill and persisted when we controlled for the prey's speed, suggesting that it would be observed even if predators hunted prey of average speed. This suggests that other proximal or ecological factors may be at play. For example, predators have limited attention spans that restrict diet choice and the formation of search images when they need to focus on hunting multiple prey simultaneously [@dukas.etal2001;@troscianko.etal2018]. Chasing prey demands specialized cognitive abilities and coordination that are energetically costly [@kelley.magurran2011]. Hence, individuals with neurological or physical attributes allowing them to better concentrate on the hunting task may have shown greater improvement in success over time [@dukas2019]. Alternatively, predators may also shape their own expertise via their hunting tactic. For example, faster hunters may push their prey to move faster and display altruistic behaviours, which, in turn, may hinder their own expertise acquisition due to the increased difficulty of hunting such prey [@fraserfranco.etal2022;@cere.etal2024]. Lastly, longer time intervals between hunting events are hypothesized to hinder or delay the acquisition of expertise due to information loss [@endler1991;@wright.etal2022;@wooster.etal2023]. For example, a predator who played 300 matches within six months might forget critical information related to prey detection or escape patterns compared to one who played 300 matches within six weeks. Therefore, investigating the impact of the aforementioned elements may reveal important insights into the outcome of predator-prey interactions, but integrating them into a unified framework remains challenging due to computational constraints and limitations in current modeling approaches.
